---
title: "The role of personality in shaping music preferences"
Author: K24091801
output:
  html_document:
    df_print: paged
---

##Introduction

Music is embedded in our daily lives, shaping how individuals feel, think and connect with the world. From casual listening to moments of deep emotional connection, music has long been recognized as a tool for managing emotions, social bonding and facilitating cultural expression. (Vuust & Kringelbach, 2010) Especially for young people, music is not just seen as a form of entertainment but as a means to self-expression. Music as a form of art, is deeply rooted in culture serving us both for individual or collective experience. Research has highlighted how music plays a role in shaping identities, as young people tend to use music to connect with social groups and express their individuality. Beyond this, music also fosters a sense of belonging or enabling shared memories, playing a critical role in social and cultural interaction. (Laughey, 2006)

Musical preferences can be therefore categorized at different levels using different approaches one of them being music genres. Dorochowicz et al. (2020) have demonstrated how specific personalities resonate better to specific genres such as introverts preferring energetic music like hip-hop while  classical and jazz music for introverts, 

 
This foundational understanding of music’s cultural and psychological dimensions raises many questions regarding how individual personality traits influence music preferences among young people. Personality traits, a key area of  psychological studies, shape individuals’ views on life and judgement. On the other hand, music is a fundamental component of culture that reflects broader societal values. Highlighting on the interplay between psychological and cultural dimensions, Albury (2024) conducted a data study on how personal and social factors influence musical taste particularly in understanding how certain variables such as gender or environment shape musical engagement. Renfrow and Levitin (2019) emphasizes the psychological basis of music preferences arguing how personality traits influence the adoption of certain musical genres that align with an individual's identity and values. 

For instance, individuals with reflective tendencies tend to prefer complex genres like jazz, the article highlights how some preferences not only reflect individual psychological traits but broader patterns of cultural engagement.
 
Some studies have suggested that demographic variables significantly influence musical engagement. Gong's (2020) explored the relationship between genre and educational level influences music preferences, finding that male students tend to prefer rock and roll compared to young females showing a preference in classical music. Therefore it is important to analyse how musical preferences not only interact with personality but if in reality demographic factors also play a role in this. Because of this,  Greasley and Lamont (2016) mentioned how music should not be studied as a static expression of taste but as a dynamic reflection of life since our environment tends to affect our opinions. 
 
Despite advancements in understanding the psychology of music significant gaps remain. Most of the existing research is mainly based on theoretical research and little quantitative is added. Moreover, there has been insufficient exploration of how demographic variables like age, gender, and education mediate this relationship. For instance, Gong (2020) and Bonneville-Roussy et al. (2013)  do address gender and age as demographic factors but little research has been done to understand how these variables interact and the correlation to it. 
 
Therefore, this research aims to explore the gaps in understanding how personality traits and demographic variables such as age, gender, environment and daily behaviours influence music preferences among young people. Using larger and more varied samples, this research aims to discover new patterns. The analysis will be of quantitative nature, therefore it will be based on integer variables. 


Do demographic variables like age, gender, and education level assossiation the relationship between personality traits and music preferences?


##Datasets 

For this project, the data source used is the Young People Survey from Kaggle (Sabo, 2017) containing a .csv file with the dataset. All survey participants were Slovakian nationality, aged between 15-30. . The data contains 1,010 observations on 150 attributes. The dataset contains 608 missing data points on 336 records, with only 674 rows with no missing values.

```{r}

library(dplyr);
library(corrplot);
library(ggplot2);
library(tidyr);

```

Loading the dataset

```{r}
data=read.csv("responses.csv");

```

Making a copy of the original dataset incase its needed in future 
```{r}
responses <- data.frame(data) ;

```


##Data Preparation




#Data Cleaning

Firstly, converted the categorical columns with categorical values into numeric values to fit in our analysis methods such as correlation. 
```{r}
data$Gender <- recode(data$Gender,"female"=1,"male"=2);

data$Alcohol <- recode(data$Alcohol,"drink a lot"=1,"social drinker"=2,"never"=3);


education <- c("currently a primary school pupil"=1, "primary school"=2, "secondary school"=3,
        "college/bachelor degree"=4,"masters degree"=5,"doctorate degree"=6);

data$Education <- recode(data$Education,!!!education);


data$Smoking <- recode(data$Smoking,"tried smoking"=2,"current smoker"=3,
                       "former smoker"=4,"never smoked"=1);

lying <- c("everytime it suits me"=1,"only to avoid hurting someone"=3,
        "sometimes"=2,"never"=4);

data$Lying <- recode(data$Lying,!!!lying);


data$Internet.usage <- recode(data$Internet.usage,"few hours a day"=3,
                              "less than an hour a day"=2,
                              "most of the day"=4,"no time at all"=1);

data$Punctuality <- recode(data$Punctuality,"i am often early"=1,"i am always on time"=2,
                           "i am often running late"=3);

data$House...block.of.flats <- recode(data$House...block.of.flats,
                                      "block of flats"=1,"house/bungalow"=2);

data$Village...town <- recode(data$Village...town,"village"=1,"city"=2);

data$Left...right.handed <- recode(data$Left...right.handed,"left handed"=1,
                                   "right handed"=2);

table(data$Only.child)

data$Only.child <- recode(data$Only.child,"yes"=1,"no"=2)

table(data$Only.child)

```
I checked... 
```{r}

table(data$Only.child)

data$Only.child <- recode(data$Only.child,"yes"=1,"no"=2)

table(data$Only.child)

```


#Handling Missing Values

Checking missing values
```{r}

sum(is.na(data));

complete_data=complete.cases(data);

dim(data[complete_data,])


```

The sum of missing values is 608

Missing values for variable - this is why i choose exclusion 
```{r}
colSums(is.na(data))
```


674 complete rows 

```{r}

data_clean <- data[complete.cases(data), ]

```

Confirming that all the missing values have been replaced
```{r}
sum(is.na(data_clean));

complete_data=complete.cases(data_clean);

dim(data_clean[complete_data,])
```

```{r}
cat("Original dimensions:", dim(data), "\n")
cat("Cleaned dimensions:", dim(data_clean), "\n")

```


#Converting 

##Data Analysis


#Exploratory Data Analysis 

1. Visualizing Music Genre Popularity

```{r}

genre<-data_clean[columns[3:19]];


column_averages <- colMeans(genre, na.rm = TRUE)

# Step 2: Sort the column averages in descending order
sorted_averages <- sort(column_averages, decreasing = TRUE)

# Step 3: Create a bar plot of the sorted averages
barplot(sorted_averages, 
        main = "Music Genre Popularity.",  # Title of the plot
        col = "lightblue",  # Bar color
        las = 2,  # Rotate axis labels for readability
        ylab = "Average",  # Label for the y-axis
        cex.names = 0.8)  # Scale the axis labels for readability
```


Generally, rock music genre is most preferred by most people. On average individuals surveyed have 4.0 preference score for the genre, asserting that the is highly loved among the young people. Pop genre follows closely with an average preference score of 3.5 while Opera and Country are least preferred, respectively. The following graph illustrates genre popularity.


2. Music genre preference score by Gender


```{r}

genre<-data_clean[columns[3:19]];

gender_genre<-data.frame(data_clean["Gender"],genre);

gender_genre <- gender_genre %>%
  group_by(Gender) %>%
  summarize(across(where(is.numeric), mean, na.rm = TRUE))

gender_genre <- gender_genre %>%
  pivot_longer(cols = -Gender, names_to = "Music_Genre", values_to = "Score")
gender_genre$Gender <- as.factor(gender_genre$Gender)

# Plot using ggplot2
ggplot(gender_genre, aes(y =Music_Genre, x =Score, fill = (Gender))) + 
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Music Genre Preference Score by Gender",
       x = "Music Genre",
       y = "Average Score") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels
  theme_minimal() +
  scale_fill_manual(values = c( "1"= "pink", "2" = "blue")) 
```



3. Music genre preference score by Age 


```{r}
age_music <- data_clean[c("Age", "Music")]

age_music <- age_music %>%
  group_by(Age) %>%
  summarize(Average = mean(Music))

# View the result
ggplot(age_music, aes(x = Age, y = Average)) + 
  geom_bar(stat = "identity", fill = "skyblue") + 
  labs(title = "Music Preference By Age", x = "Age", y = "Average Preference for Music") +
  theme_minimal()

genre<-data_clean[columns[3:19]];

age_genre<-data.frame(data_clean["Age"],genre);

age_genre <- age_genre %>%
  group_by(Age) %>%
  summarize(across(where(is.numeric), mean, na.rm = TRUE))

age_genre <- age_genre %>%
  pivot_longer(cols = -Age, names_to = "Music_Genre", values_to = "Score")

# Plot using ggplot2
ggplot(age_genre, aes(y = Music_Genre, x = Score, fill = as.factor(Age))) + 
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Music Genre Preference Score by Age",
       x = "Music Genre",
       y = "Average Score") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels
  theme_minimal()
```
```





```{r}

age_music <- data_clean[c("Age", "Music")]

age_music <- age_music %>%
  group_by(Age) %>%
  summarize(Average = mean(Music))


# View the result
ggplot(age_music, aes(x = Age, y = Average)) + 
  geom_bar(stat = "identity", fill = "skyblue") + 
  labs(title = "Music Preference By Age", x = "Age", y = "Average Preference for Music") +
  theme_minimal()


genre<-data_clean[columns[3:19]];

age_genre<-data.frame(data_clean["Age"],genre);

age_genre <- age_genre %>%
  group_by(Age) %>%
  summarize(across(where(is.numeric), mean, na.rm = TRUE))

age_genre <- age_genre %>%
  pivot_longer(cols = -Age, names_to = "Music_Genre", values_to = "Score")

# Plot using ggplot2
ggplot(age_genre, aes(y = Music_Genre, x = Score, fill = as.factor(Age))) + 
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Music Genre Preference Score by Age",
       x = "Music Genre",
       y = "Average Score") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels
  theme_minimal()
```


```{r}

```



```{r}


mean 

BAR PLOT FREQUENCY
```{r}

age_rock=age_genre[age_genre$Music_Genre=="Rock",]

barplot(age_rock$Age, age_rock$Score)



```




4. Music genre preference score by environment 









#Correlation Analysis 


1. Correlation between demographic information and music preference

The graph below shows how demographic information such age, education, gender , and type of residence correlate with individuals music taste.

threshold 0.5 

```{r}
columns <- c(colnames(data_clean));

#make a subset of the demographic columns
demographics <- data_clean[columns[77:133]]
music<-data_clean[columns[3:10]]

#Combining Music and demographic subsets

music_demo=data.frame(music, demographics)

corrplot(cor(music_demo));
```



vedi la suddivisione interna 

```{r}

columns <- c(colnames(data_clean));

#make a subset of the demographic columns
traits <- data_clean[columns[77:133]]
music<-data_clean[columns[3:10]]

#Combining Music and demographic subsets

music_demo=data.frame(music, traits)

corrplot(cor(music_demo));


```


From the correlation plot above, there appears to be minimal relationship between person’s  demographic information with respect to their music preferences. However , there is notables relationship between ones’  music preference with where they spend most of their childhood in. Particularly, individuals who spent most of their childhood in the village tend to prefer folk genre. Additionally, individuals who spent most of their childhood in houses/bungalow tend to prefer folk music genre.




2. Correlation between personal traits with music preference (SPEARMAN - METTERE UN RANK TRA I DATI )




SPEARMAN





```{r}

# Create a list to store the person_trait subsets
person_traits <- list()

# Loop through and create subsets from columns 77 to 133, with intervals of 5
for (i in seq(77, 129, by = 10)) {
  end_index <- min(i + 4, 133)  # Ensure we don't go beyond column 133
  person_traits[[length(person_traits) + 1]] <- data_clean[columns[i:end_index]]
}

# Now create music_trait datasets and plot correlations dynamically
for (i in 1:length(person_traits)) {
  music_trait <- data.frame(music, person_traits[[i]])
  corrplot(cor(music_trait))
}

```

There is minimal relationship between music preference and person’s traits. Workaholic individuals appears to prefer classical music as opposed to the other music genres. Similary, individuals who used to cheat in school prefer hip-pop and rap music. People who usually give to charity prefer opera while those who believe in God prefer folk genres



##Conclusion and Limitation

how i handle missing values 



##References



#End
